\chapter{Background}
\label{chapter:Background}

The key abstraction of functional programming is the function, which takes an
input and produces an output. Functions may produce functions as output and/or
take functions as input. We say that functions are {\em first-class values} in a
functional programming language.

In contrast, the more popular imperative model of programming takes statements,
or actions which modify the state of the world, as the primary building blocks
of programs. Such programs have sequential control flow, and require reasoning
about side-effects. They are thus inherently resistant to compositional
reasoning.

Even in functional languages, reactive programs are generally written in an imperative style, using low-level and non-composable abstractions including callbacks
or object-based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models. 

{\em Functional} Reactive Programming implies that a model should keep the characteristics of functional programming (i.e. that the basic constructs of the language
should remain first-class) while incorporating reactivity into the language model. In particular, functions should be lifted to operate on reactive values,
and functions themselves ought to be reactive.

The goal of FRP is to provide compositional and high-level abstractions for
creating reactive programs. The key abstractions of FRP are behaviors
\footnote{Behaviors are generally referred to as {\em signals} in
signal-function literature. This is unfortunate, since a {\em signal function}
may operate on events, behaviors, or some combination of the two.}, which are
time-dependent values defined at every point in continuous time, and events,
which are values defined at countably many points in time. An FRP system will
provide combinators to manipulate events and behaviors and to react to events by
replacing a portion of the running program in response to an event. Behaviors
and events, or some abstraction based on them, will be first class, in keeping
with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable, but this has proven to be the main
challenge in implementing FRP.

The two general approaches to FRP are ``classic'' FRP, where behaviors and
events are first-class and reactive objects, and ``signal-function'' FRP, where
transformers of behaviors and events are first-class and reactive objects.

\section{Classic FRP}
\label{section:Background-Classic_FRP}

The earliest and still standard formulation of FRP provides two primitive type
constructors: {\tt Behavior} and {\tt Event}, together with combinators that
produce values of these types. The easiest semantic definition for these types
is given in Figure~\ref{figure:classic_frp_semantic_types}\footnote{The list type constructor {\tt []} should be considered to contain infinite as well as finite lists.}.

\begin{figure}
\begin{code}
type Event a = [(Time, a)]
type Behavior a = Time -> a
\end{code}
\hrule
\caption{Semantic types for Classic FRP.}
\label{figure:classic_frp_semantic_types}
\end{figure}


When these two type constructors are exposed directly, the system is known as a 
{\em Classic FRP} system. If the type aliases are taken as given in the semantic
definition, a simple, yet problematic, implementation is given in
Figure~\ref{figure:classic_frp_simple_implementation}.

\begin{figure}
\begin{code}
time :: Behavior Time
time = id

constant :: a -> Behavior a
constant = const

delayB :: a -> Time -> Behavior a -> Behavior a
delayB a td b te = if te <= td
                   then a
                   else b (te - td)          

instance Functor Behavior where
  fmap = (.)

instance Applicative Behavior where
  pure = constant
  bf <*> ba = (\t -> bf t (ba t))

never :: Event a
never = []

once :: a -> Event a
once a = [(0, a)]

delayE :: Time -> Event a -> Event a
delayE td = map (\(t, a) -> (t + td, a))

instance Functor Event where
  fmap f = map (\(t, a) -> (t, f a))

switcher :: Behavior a -> Event (Behavior a) -> Behavior a
switcher b [] t = b t
switcher b ((to, bo):_) t = if t < to
                            then b t
                            else bo t
\end{code}
\hrule
\caption{An obvious, yet inefficient and problematic, implementation of
Classic FRP.}
\label{figure:classic_frp_simple_implementation}
\end{figure}

There are several obvious problems with this implementation of course, but it
suffices to show the intuition behind the classic FRP model. Problems in this
implementation which are addressed by real implementations include the necessity
of waiting for event occurrences, the necessity of maintaining the full history
of a behavior in memory, and the lack of an enforced order for event
occurrences.

\subsection{History of Classic FRP}
\label{subsection:Background-Classic_FRP-History_of_Classic_FRP}
Classic FRP was originally described as the basis of Fran~\cite{Elliott1997}
(Functional Reactive ANimation), a framework for declaratively specifying
interactive animations. Fran represented behaviors as two sampling functions,
one from a time to a value and a new behavior (so that history may be discarded),
and the other from a time interval (a lower and upper time bound) to a value
interval and a new behavior. Events are represented as ``improving values'',
which, when sampled with a time, produce a lower bound on the time to the next
occurrence, or the next occurrence if it has indeed occurred.

The first implementation of FRP outside the Haskell language was Frapp\'{e}~\cite{Courtney2001-2},
implemented in the Java Beans framework. Frapp\'{e} built on the notion
of events\footnote{These are semantic objects to which callbacks may be added,
and not events in the FRP sense, though they are related.} and ``bound
properties'' in the Beans framework, providing abstract interfaces for FRP
events and behaviors, and combinators as concrete classes implementing these
interfaces. The evaluation of Frapp\'{e} used Bean components as sources and
sinks, and the implementation of Bean events and bound properties to propagate
changes to the network.

\subsection{Current Classic FRP Systems}
\label{subsection:Background-Classic_FRP-Current_Classic_FRP_Systems}

The FrTime\footnote{FrTime is available in all recent versions of Dr. Racket}~cite{Cooper2006}
language extends the Scheme evaluator with a mutable dependency graph, which is
constructed by program evaluation. This graph is then updated by signal changes.
FrTime does not provide a distinct concept of events, and chooses branches of
the dependency graph by conditional evaluation of behaviors, rather than by the
substitution of behaviors used by FRP systems.

The Reactive~\cite{Elliott2009} system is a push-pull FRP system with
first-class behaviors and events. The primary insight of Reactive is the
separation of reactivity (or changes in response to events whose occurrence time
could not be known beforehand) from time-dependence. This gives rise to {\em
reactive normal form}, which represents a behavior as a constant or simply
time-dependent value, together with an event stream carrying values which are also
behaviors in reactive normal form. Push-based evaluation is achieved by forking
a Haskell thread to evaluate the head behavior, while waiting on the evaluation
of the event stream. Upon an event occurrence, the current behavior thread is
killed and a new thread spawned to evaluate the new behavior. Unfortunately, the
implementation of Reactive uses a tenuous technique which depends on forking
threads to evaluate two Haskell values concurrently, in order to implement event
merging. This relies on the library author to ensure consistency when this
technique is used, and leads to thread leakage when one of the merged events is
itself the merging of other events.

A recent thesis~\cite{Czaplicki2012} described Elm, a stand-alone language for
reactivity. Elm provides combinators for manipulating discrete events, and
compiles to JavaScript, making it useful for client-side web programming.
However, Elm does not provide a notion of switching or continuous time
behaviors, though an approximation is given using discrete-time events which are
actuated at repeated intervals specified during the event definition. The thesis
asserts that Arrowized FRP (signal-function FRP, Section~\ref{section:Background-signal_function_frp})
can be embedded in Elm, but provides little support for this assertion\footnote{
A form of Arrowized FRP employing applicative functors is presented, and
justified by the assertion that applicative functors are just arrows with the
input type parameter fixed. While this is true, it ignores the compositional
benefits of the general Arrow framework, and it is not clear that it provides
the same enforcement of causality.}

The reactive-banana~\cite{Apfelmus} library is a push-based FRP system designed
for use with Haskell GUI frameworks. In particular, it features a monad for
the creation of behaviors and events which may then be composed and evaluated.
This monad includes constructs for binding GUI library constructs to primitive
events. It must be ``compiled'' to a Haskell IO action for evaluation to take
place. The implementation of reactive-banana is similar to FrTime, using a
dependency graph to update the network on event occurences. Reactive-banana
eschews generalized switching in favor of branching functions on behavior
values, similarly to FrTime, but maintains the distinction between behaviors and
events. Rather than a generalized switching combinator which allows the
replacement of arbitrary behaviors, reactive-banana provides a step combinator
which creates a stepwise behavior from the values of an event stream.

\section{Signal Function FRP}
\label{section:Background-signal_function_frp}

An alternative approach to FRP was first proposed in work on
Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs.
This library utilized the concept of Arrows~\cite{Hughes2000} as an abstraction
for {\em signal functions}. Arrows are abstract type constructors with input and
output type parameters, together with a set of routing combinators (see
Figure~\ref{figure:arrow_combinators}). The concept of an Arrow in Haskell,
including the axioms the combinators of an arrow must satisfy, are derived from
the concept of arrows from category theory.

Signal functions are time-dependent and reactive transformers of events and
behaviors. Behaviors and events cannot be directly manipulated by the programmer.
This approach has two motivations: it increases modularity since both the input
and output of signal functions may be transformed (as opposed to behaviors or
events which may only be transformed in their output) and it avoids a large
class of time and space leaks which have emerged when implementing FRP with
first-class behaviors and events.

\begin{figure}
\begin{code}
(>>>)  :: (Arrow a) => a b c -> a c d -> a b d
arr    :: (Arrow a) => (b -> c) -> a b c
first  :: (Arrow a) => a b c -> a (b, d) (c, d)
second :: (Arrow a) => a b c -> a (d, b) (d, c)
(***)  :: (Arrow a) => a b c -> a b d -> a b (c, d)
loop   :: (Arrow a) => a (b, d) (c, d) => a b c
\end{code}
\hrule
\caption{Arrow combinators.}
\label{figure:arrow_combinators}
\end{figure}

Similarly to FrTime, the netwire~\cite{Soylemez} library eschews dynamic
switching, in this case in favor of {\em signal inhibition}. Netwire is written
as an arrow transformer, and permits the lifting of IO actions as sources and
sinks at arbitrary points in a signal function network. Signal inhibition is
accomplished by making the output of signal functions a monoid, and then
combining the outputs of signal functions. An inhibited signal function will
produce the monoid's zero as an output. Primitives have defined inhibition
behavior, and composed signal functions inhibit if their outputs combine to the
monoid's zero.

Yampa~\cite{Nilsson2005} is an optimization of the Arrowized FRP system first
utilized with Fruit (see above). The implementation of Yampa makes use of
Generalized Algebraic Datatypes to permit a much larger class of type-safe
datatypes for the signal function representation. This representation,
together with ``smart'' constructors and combinators, enables the construction
of a self-optimizing arrowized FRP system. Unfortunately, the primary
inefficiency, that of unnecessary evaluation steps due to pull-based evaluation,
remains. Further, the optimization is ad-hoc and each new optimization requires
the addition of new constructors, as well as the updating of every primitive
combinator to handle every combination of constructors. However, Yampa showed
noticeable performance gains over previous Arrowized FRP implementations.

A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for
typing Arrowized FRP systems using dependent types. The bulk of the work
consisted in using the dependent type system to prove the correctness of the FRP
system discussed. This work introduced signal vectors, a typing construct
which permits the distinction of behavior and event types at the level of the FRP
system, instead of making events merely a special type of behavior.

\section{Outstanding Challenges}
\label{section:Background-outstanding_challenges}

At present, there are two key issues apparent with FRP. First, while
signal-function FRP is inherently safer and more modular than classic FRP, it
has yet to be efficiently implemented. Classic FRP programs are vulnerable to
time leaks and violations of causality due to the ability to directly manipulate
reactive values. Second, the interface between FRP programs and the many
different sources of inputs and sinks for outputs available to a modern
application writer remains ad-hoc and is in most cases limited by the
implementation.

One key exception to this is the reactive-banana system, which provides a monad
for constructing primitive events and behaviors from which an FRP program may
then be constructed. However, this approach is still inflexible as it requires
library support for the system which the FRP program will interact with.
Further, being a Classic FRP system, reactive-banana lacks the ability to
transform the inputs of behaviors and events, since all inputs are implicit.
