\documentclass[letterpaper]{article}
\usepackage[margin=1.2in]{geometry}
\usepackage{ifthen}
\usepackage{stmaryrd}
\usepackage{mathpartir}
\usepackage[usenames]{color}
\usepackage{hyperref}
\usepackage{multicol}
\usepackage{fancyvrb}

% For HsColour
%\newsavebox{\spaceb}
%\newsavebox{\tabb}
%\savebox{\spaceb}[1ex]{~}
%\savebox{\tabb}[4ex]{~}
%\newcommand{\hsspace}{\usebox{\spaceb}}
%\newcommand{\hstab}{\usebox{\tabb}}
%\newcommand{\conceal}[1]{}

\input{pygments.tex}
\input{amsden-commands.tex}


\title{Push-Based Signal-Function Functional Reactive Programming (Thesis Proposal)}
\author{Edward Amsden}
\date{\today}

\begin{document}

\maketitle
\input{abstract.tex}
\input{introduction.tex}
\input{background.tex}
\input{proposed_work.tex}

\bibliographystyle{acm}
%\bibliography{proposal}
\begin{thebibliography}{10}

\bibitem{Akeley2011}
{\sc Akeley, K., and Segal, M.}
\newblock {\em The OpenGL Graphics System: A Specification}, 2011.

\bibitem{Apfelmus}
{\sc Apfelmus, H.}
\newblock reactive-banana library.
\newblock \mbox{\url{http://hackage.haskell.org/package/reactive-banana}}.

\bibitem{Cooper2006}
{\sc Cooper, G.~H., and Krishnamurthi, S.}
\newblock Embedding dynamic dataflow in a call-by-value language.
\newblock In {\em European Symposium on Programming\/} (2006), pp.~294--308.

\bibitem{Courtney2001-2}
{\sc Courtney, A.}
\newblock Frapp\'{e}: Functional reactive programming in {J}ava.
\newblock In {\em Proceedings of the Third International Symposium on Practical
  Aspects of Declarative Languages\/} (London, UK, UK, 2001), PADL '01,
  Springer-Verlag, pp.~29--44.

\bibitem{Courtney2001-1}
{\sc Courtney, A., and Elliott, C.}
\newblock Genuinely functional user interfaces.
\newblock In {\em In Proceedings of the 2001 Haskell Workshop\/} (2001),
  pp.~41--69.

\bibitem{Czaplicki2012}
{\sc Czaplicki, E.}
\newblock Elm: Concurrent {FRP} for functional {GUI}s.
\newblock
  \url{http://www.testblogpleaseignore.com/wp-content/uploads/2012/03/thesis.pdf},
  2012.

\bibitem{Elliott2009}
{\sc Elliott, C.}
\newblock Push-pull functional reactive programming.
\newblock In {\em Haskell Symposium\/} (2009).

\bibitem{Elliott1997}
{\sc Elliott, C., and Hudak, P.}
\newblock Functional reactive animation.
\newblock In {\em Proceedings of the second ACM SIGPLAN international
  conference on Functional programming\/} (New York, NY, USA, 1997), ICFP '97,
  ACM, pp.~263--273.

\bibitem{Hughes2000}
{\sc Hughes, J.}
\newblock Generalising monads to arrows.
\newblock {\em Science of Computer Programming 37}, 1–3 (2000), 67 -- 111.

\bibitem{Kilgard96}
{\sc Kilgard, M.~J.}
\newblock {\em The OpenGL Utility Toolkit (GLUT) Programming Interface API
  Version 3}, 1996.

\bibitem{Nilsson2005}
{\sc Nilsson, H.}
\newblock Dynamic optimization for functional reactive programming using
  generalized algebraic data types.
\newblock In {\em Proceedings of the tenth ACM SIGPLAN international conference
  on Functional programming\/} (New York, NY, USA, 2005), ICFP '05, ACM,
  pp.~54--65.

\bibitem{Nilsson2002}
{\sc Nilsson, H., Courtney, A., and Peterson, J.}
\newblock Functional reactive programming, continued.
\newblock In {\em Proceedings of the 2002 ACM SIGPLAN workshop on Haskell\/}
  (New York, NY, USA, 2002), Haskell '02, ACM, pp.~51--64.

\bibitem{Paterson2001}
{\sc Paterson, R.}
\newblock A new notation for arrows.
\newblock In {\em Proceedings of the sixth ACM SIGPLAN international conference
  on Functional programming\/} (New York, NY, USA, 2001), ICFP '01, ACM,
  pp.~229--240.

\bibitem{Sculthorpe2011}
{\sc Schulthorpe, N.}
\newblock {\em Towards Safe and Efficient Functional Reactive Programming}.
\newblock PhD thesis, University of Nottingham, UK, 2011.

\bibitem{Soylemez}
{\sc S\"{o}ylemez, E.}
\newblock netwire library.
\newblock \mbox{\url{http://hackage.haskell.org/package/netwire-3.1.0}}.

\end{thebibliography}



\end{document}